Data Exploration: IUCN and AquaMaps

### 2015-11-18 data_explore.Rmd
uploading new figures etc

### Inspect each quadrant to find representative species
# quadrant1_spp <- spp_list %>%
#   filter(area_ratio > 90 & sm_perc > 90)
# quadrant2_spp <- spp_list %>%
#   filter(area_ratio > 25 & area_ratio < 30) %>%
#   filter(sm_perc > 90)
# quadrant4_spp <- spp_list %>%
#   filter(sm_perc > 25 & sm_perc < 30) %>%
#   filter(area_ratio > 90)
# quadrant3_spp <- spp_list %>%
#   filter(area_ratio < 25 & sm_perc < 25)

spp_quads <- c('q1' = 'Kajikia albida', 
               'q2' = 'Galaxea cryptoramosa', 
               'q3' = 'Acanthurus nigroris', 
               'q3a' = 'Metanephrops australiensis', 
               'q4' = 'Conus magnificus')

@@

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(area_ratio, sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  geom_point(data = spp_list_label, color = 'grey30') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'grey30',
            fontface = 3,
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

@@

@@

spp <- fread(input = file.path(dir_aquamaps, 'tables/ohi_speciesoccursum.csv'), 
             verbose = FALSE, header = TRUE) %>%
  mutate(sciname = paste(Genus, Species, sep = ' '))
spp <- as.data.frame(spp)

@@

```{r, warning = FALSE, message = FALSE, echo = FALSE}

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  mutate(prop = cells_0.4 / cells_t) %>%
  left_join(spp, by = c('species_id' = 'SPECIESID')) %>%
  select(sciname, species_id, cells_0.4, cells_t, prop, iucn_code)

@@

>>>>

area_ratio_quartiles <- quantile(spp_list$area_ratio, na.rm = TRUE)[2:4]
pct_overlap_quartiles <- quantile(spp_list$sm_perc, na.rm = TRUE)[2:4]

spp_quads <- c('q1' = 'Kajikia albida', 
               'q2' = 'Galaxea cryptoramosa', 
               'q4' = 'Acanthurus nigroris', 
               'q4a' = 'Metanephrops australiensis', 
               'q3' = 'Conus magnificus')

@@

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = log_total_area)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  geom_vline(x = area_ratio_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_point(data = spp_list_label, color = 'tomato2') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'tomato2',
            fontface = 'bold.italic',
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

@@

scatter_arearatio_vs_overlap2 <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  geom_vline(x = area_ratio_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_point(data = spp_list_label, color = 'grey30') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'grey30',
            fontface = 'bold.italic',
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

ggsave('figures/AreaRatiovsPercOverlap2.png', width = 6, height = 4)

print(scatter_arearatio_vs_overlap2)

``` {r examine_spp_gp_quartiles}

spp_list1 <- spp_list %>%
  group_by(spp_group) %>%
  mutate(n_spp = n()) %>%
  ungroup()

spp_gp_q1 <- spp_list1 %>%
  filter(area_ratio >= median(area_ratio, na.rm = TRUE) & sm_perc >= median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q1 = n())
spp_gp_q2 <- spp_list1 %>%
  filter(area_ratio < median(area_ratio, na.rm = TRUE) & sm_perc >= median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q2 = n())
spp_gp_q4 <- spp_list1 %>%
  filter(area_ratio < median(area_ratio, na.rm = TRUE) & sm_perc < median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q4 = n())
spp_gp_q3 <- spp_list1 %>%
  filter(area_ratio >= median(area_ratio, na.rm = TRUE) & sm_perc < median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q3 = n())

spp_gp_quadrants <- spp_gp_q1 %>%
  left_join(spp_gp_q2, by = c('spp_group', 'n_spp')) %>%
  left_join(spp_gp_q3, by = c('spp_group', 'n_spp')) %>%
  left_join(spp_gp_q4, by = c('spp_group', 'n_spp')) %>%
  gather(quad, n_quad, n_spp_q1, n_spp_q2, n_spp_q3, n_spp_q4) %>% 
  mutate(quad = str_replace(quad, 'n_spp_', ''),
         pct_quad = n_quad/n_spp)

spp_gp_quadrants <- spp_gp_quadrants %>%
  left_join(spp_gp_quadrants %>% 
              filter(quad == 'q1') %>%
              select(spp_group, pct_q1 = pct_quad),
            by = 'spp_group')

barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group = reorder(spp_group, pct_q1)), 
                                aes(x = spp_group, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin') +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0) +
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Relative number of species per quadrant', 
       title = 'Species group breakdown by quadrant')

@@

spp <- fread(input = file.path(dir_aquamaps, 'tables/ohi_speciesoccursum.csv'), 
             verbose = FALSE, header = TRUE) %>%
  mutate(sciname = paste(Genus, Species, sep = ' '),
         iucn_code = str_replace(toupper(iucn_code), 'LR/', ''),
         iucn_code = str_replace(iucn_code, 'N.E.', 'NE'))
spp <- as.data.frame(spp)

@@

```{r hist_lost_area_w_threshold, warning = FALSE, message = FALSE, echo = FALSE}

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  mutate(prop = cells_0.4 / cells_t) %>%
  left_join(spp %>%
#              filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')), 
              filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX')), 
            by = c('species_id' = 'SPECIESID')) %>%
  select(sciname, species_id, cells_0.4, cells_t, prop, iucn_code)

@@

```{r scatter_lost_area_w_threshold, warning = FALSE, message = FALSE, echo = FALSE}

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat

threshold_df <- data.frame()
for (i in seq(0, 1, .05)) { # i = .05
  temp_df <- spp_cells %>%
    group_by(species_id) %>%
    summarize(thresh       = i,
              cells_thresh = sum(probability >= thresh),
              cells_total  = sum(probability >= 0),
              proportion   = cells_thresh/cells_total) %>%
    left_join(spp %>%
#              filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')), 
              filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX')), 
            by = c('species_id' = 'SPECIESID')) %>%
    select(sciname, species_id, thresh, cells_thresh, cells_total, proportion, iucn_code)

  threshold_df <- threshold_df %>%
    bind_rows(temp_df)
}

thresh_df_summary <- threshold_df %>%
  group_by(thresh) %>%
  summarize(mean_prop   = mean(proportion, na.rm = TRUE),
            median_prop = median(proportion, na.rm = TRUE))

scatter_area_vs_threshold <- ggplot(thresh_df_summary, aes(x = thresh, y = mean_prop)) + 
  ggtheme_plot + 
  geom_point(aes(y = median_prop), color = 'grey60') +
  geom_point() + 
  geom_abline(slope = -1, intercept = 1, color = 'red') + 
  geom_vline(x = 0.40, color = 'blue', linetype = 'dashed') +
  ylim(0, 1) +
  labs(title = 'Mean (median) species range after presence threshold cutoff',
       x = 'Presence threshold cutoff', 
       y = 'Mean (median) relative species range')

ggsave('figures/scatter_AMThreshold.png', width = 6, height = 4)

print(scatter_area_vs_threshold)
